---
title: "Causal Inference in Observational Studies: A Doubly Robust Approach"
author:
  - name: Alice Chen
    affiliations:
      - Department of Biostatistics, University of Example
  - name: Bob Williams
    affiliations:
      - Department of Epidemiology, University of Example
  - name: Diana Park
    affiliations:
      - Department of Statistics, University of Example
date: 2025-06-15
format: pdf
bibliography: references.bib
abstract: |
  Causal inference from observational data remains a central challenge
  in epidemiology. We compare propensity score methods with doubly
  robust estimators and demonstrate their application using data from
  the National Health Survey and the BRFSS.
---

## Introduction

Establishing causal relationships from observational data is a
fundamental challenge in public health research [@hernan_causal_2020;
@robins_new_1986]. Unlike randomized trials, observational studies are
susceptible to confounding, which can bias estimated treatment effects.

Propensity score methods [@rosenbaum_central_1983] have become a
standard tool for addressing confounding in observational research.
These methods model the probability of treatment assignment given
observed covariates, enabling researchers to create balance between
treatment groups. More recently, doubly robust estimators
[@bang_doubly_2005] have gained popularity as they provide consistent
estimates if either the propensity score model or the outcome model is
correctly specified. See @sec-methods for our analytical approach and
@sec-results for findings.

The average treatment effect (ATE) is defined as $\text{ATE} = E[Y(1) - Y(0)]$,
where $Y(1)$ and $Y(0)$ represent potential outcomes under treatment
and control, respectively. Under the doubly robust framework, the
augmented estimator takes the form:

$$
\hat{\tau}_{DR} = \frac{1}{n} \sum_{i=1}^{n} \left[ \frac{A_i Y_i}{\hat{e}(X_i)} - \frac{(A_i - \hat{e}(X_i)) \hat{m}_1(X_i)}{\hat{e}(X_i)} \right] - \frac{1}{n} \sum_{i=1}^{n} \left[ \frac{(1-A_i) Y_i}{1 - \hat{e}(X_i)} + \frac{(A_i - \hat{e}(X_i)) \hat{m}_0(X_i)}{1 - \hat{e}(X_i)} \right]
$$

where $\hat{e}(X_i)$ is the estimated propensity score and
$\hat{m}_a(X_i)$ are outcome model predictions.

## Methods {#sec-methods}

### Data Source

We used data from the 2023 National Health Survey and the 2023
Behavioral Risk Factor Surveillance System (BRFSS), yielding a
combined sample of `n = 12,500` participants with complete covariate
information.

### Propensity Score Estimation

We estimated propensity scores using gradient-boosted models with
cross-validation:

$$
\hat{e}(X_i) = P(A_i = 1 \mid X_i) = f_{GBM}(X_i)
$$

where $A$ denotes treatment assignment and $X_i$ includes age, sex,
income, education, BMI, and smoking status. We assessed balance using
standardized mean differences, targeting a threshold of $d < 0.1$.

```{r}
#| label: fig-balance
#| fig-cap: "Standardized mean differences before and after propensity score weighting across both data sources."

library(ggplot2)
library(dplyr)

balance_data <- tibble(
  covariate = rep(c("Age", "Sex", "Income", "Education", "BMI", "Smoking"), 2),
  smd = c(0.35, 0.22, 0.41, 0.28, 0.19, 0.31, 0.03, 0.02, 0.04, 0.01, 0.03, 0.05),
  type = rep(c("Unadjusted", "PS-weighted"), each = 6)
)

ggplot(balance_data, aes(x = smd, y = covariate, color = type, shape = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed") +
  labs(x = "Standardized Mean Difference", y = NULL, color = "Method", shape = "Method") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Statistical Analysis

We estimated the ATE using three approaches: IPTW, outcome regression,
and the doubly robust augmented IPTW (AIPTW) estimator. Variance was
estimated using the nonparametric bootstrap with 1,000 replicates.

```{r}
#| label: tbl-model
#| tbl-cap: "Treatment effect estimates from IPTW, outcome regression, and doubly robust analysis."

library(gt)

results <- tibble(
  Model = c("Unadjusted", "IPTW", "Outcome regression", "AIPTW (doubly robust)"),
  Estimate = c(2.3, 1.6, 1.5, 1.5),
  CI_lower = c(1.5, 1.0, 0.9, 1.0),
  CI_upper = c(3.1, 2.2, 2.1, 2.0),
  p_value = c(0.001, 0.004, 0.005, 0.003)
)

gt(results) |>
  fmt_number(columns = Estimate:CI_upper, decimals = 1) |>
  tab_footnote("Bootstrap confidence intervals based on 1,000 replicates.")
```

## Results {#sec-results}

After propensity score weighting, all standardized mean differences
were below 0.1, indicating adequate balance (see @fig-balance). The
IPTW estimate of the ATE was 1.6 (95% CI: 1.0 to 2.2, p = 0.004).
The doubly robust AIPTW estimate was 1.5 (95% CI: 1.0 to 2.0,
p = 0.003), as shown in @tbl-model.

Sensitivity analyses using trimmed weights (excluding extreme
propensity scores below 0.05 or above 0.95) yielded similar results
(ATE = 1.5, 95% CI: 0.9 to 2.1). An E-value analysis indicated that
an unmeasured confounder would need to have a risk ratio of at least
2.1 with both treatment and outcome to explain away the observed
effect.

## Discussion

Our analysis demonstrates that both propensity score weighting and
doubly robust estimation effectively reduce confounding bias in this
multi-source observational dataset. The doubly robust estimate of
1.5 units is clinically meaningful and provides additional protection
against model misspecification.

These findings align with previous work by @hernan_causal_2020, who
emphasized the importance of careful confounder adjustment in
observational studies. The consistency between IPTW and AIPTW estimates
provides reassurance about the robustness of our findings. However,
neither approach can address unmeasured confounding, which remains a
key limitation. The E-value analysis provides a quantitative assessment
of this vulnerability.

## References
