---
title: "Causal Inference in Observational Studies"
author:
  - name: Alice Chen
    affiliations:
      - Department of Biostatistics, University of Example
  - name: Bob Williams
    affiliations:
      - Department of Epidemiology, University of Example
date: 2025-06-15
format: pdf
bibliography: references.bib
abstract: |
  Causal inference from observational data remains a central challenge
  in epidemiology. We review propensity score methods and demonstrate
  their application using data from the National Health Survey.
---

## Introduction

Establishing causal relationships from observational data is a
fundamental challenge in public health research [@hernan_causal_2020;
@robins_new_1986]. Unlike randomized trials, observational studies are
susceptible to confounding, which can bias estimated treatment effects.

Propensity score methods [@rosenbaum_central_1983] have become a
standard tool for addressing confounding in observational research.
These methods model the probability of treatment assignment given
observed covariates, enabling researchers to create balance between
treatment groups. See @sec-methods for our analytical approach and
@sec-results for findings.

The average treatment effect (ATE) is defined as $\text{ATE} = E[Y(1) - Y(0)]$,
where $Y(1)$ and $Y(0)$ represent potential outcomes under treatment
and control, respectively.

## Methods {#sec-methods}

### Data Source

We used data from the 2023 National Health Survey, which includes
`n = 5,000` participants with complete covariate information.

### Propensity Score Estimation

We estimated propensity scores using logistic regression:

$$
\text{logit}(P(A = 1 \mid X)) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3
$$

where $A$ denotes treatment assignment and $X_1, X_2, X_3$ are
baseline covariates. We assessed balance using standardized mean
differences, targeting a threshold of $d < 0.1$.

```{r}
#| label: fig-balance
#| fig-cap: "Standardized mean differences before and after propensity score weighting."

library(ggplot2)
library(dplyr)

balance_data <- tibble(
  covariate = rep(c("Age", "Sex", "Income", "Education"), 2),
  smd = c(0.35, 0.22, 0.41, 0.28, 0.04, 0.03, 0.06, 0.02),
  type = rep(c("Unadjusted", "PS-weighted"), each = 4)
)

ggplot(balance_data, aes(x = smd, y = covariate, color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed") +
  theme_minimal()
```

### Statistical Analysis

We estimated the ATE using inverse probability of treatment weighting
(IPTW). Variance was estimated using the robust sandwich estimator.

```{r}
#| label: tbl-model
#| tbl-cap: "Treatment effect estimates from IPTW analysis."

library(gt)

results <- tibble(
  Model = c("Unadjusted", "IPTW", "IPTW + trimming"),
  Estimate = c(2.3, 1.8, 1.7),
  CI_lower = c(1.5, 1.1, 1.0),
  CI_upper = c(3.1, 2.5, 2.4),
  p_value = c(0.001, 0.003, 0.004)
)

gt(results) |>
  fmt_number(columns = Estimate:CI_upper, decimals = 1)
```

## Results {#sec-results}

After propensity score weighting, all standardized mean differences
were below 0.1, indicating adequate balance (see @fig-balance). The
IPTW estimate of the ATE was 1.8 (95% CI: 1.1 to 2.5, p = 0.003),
as shown in @tbl-model.

Sensitivity analyses using trimmed weights (excluding extreme
propensity scores below 0.05 or above 0.95) yielded similar results
(ATE = 1.7, 95% CI: 1.0 to 2.4).

## Discussion

Our analysis demonstrates that propensity score weighting effectively
reduces confounding bias in this observational dataset. The estimated
treatment effect of 1.8 units is clinically meaningful and robust to
sensitivity analyses.

These findings align with previous work by @hernan_causal_2020, who
emphasized the importance of careful confounder adjustment in
observational studies. However, propensity score methods cannot
address unmeasured confounding, which remains a key limitation.

## References
